Repeatability of Pre-Flashover Fire Patterns on Gypsum Wallboard

### 2017-08-31 Projects/Madrzykowski_Thesis/Report/Madrzykowski_Thesis.tex
Annotate Figures, correct ref save final revision

In the United States, the loss of life due to unwanted fire has been decreasing steadily since the 1970’s.  (This trend does not include loss of life as a result of the terrorist attack on September 11, 2001.)  The estimated number of fire deaths in the U.S. in 1975 was 8,100, or 3.75 deaths per 100,000 people~\cite{America_Burning_Revisited}.  In 2013, there were 3,240 civilian fire fatalities, or 1.02 deaths per 100,000 people~\cite{Karter:2014}.  During this same period, the number of fires in the United States has decreased from 2.5 million fires to 1.2 million fires.  There are many reasons that led to this decrease in loss of life due to fire in the United States.  One example is research and data that enabled the development of product safety standards, improved building codes, improved fire safety systems, and fire prevention programs.  None of this happened without the constant effort of agencies at the federal, state, and local levels working with professional fire organizations in order to deliver the five E’s of fire prevention: engineering, education, enforcement, economic incentive, and emergency response~\cite{FEMA:2013}.

Fire investigation is a key element in the fire prevention system that generates data to enable the reduction of fire losses.  The investigation of fires provides a means to identify the cause of the fire in order to develop a knowledge base that could enable the elimination of that cause, and thus further reduce the losses from fires.  Data such as the room of fire origin, the first item ignited, and ultimately the cause of the fire is information that is critical to understanding and reducing the number of fires.  For example, the identification of products that are involved in ignitions due to a design flaw in many cases result in the U.S. Consumer Product Safety Commission or a manufacturer issuing a product recall.

@@

\subsection{The Use of Fire Models in Fire Investigation}
Fire models are another tool that may be used by a fire investigation team to gain insight into the growth and development of a fire and to test the investigators' hypothesis on origin and cause~\cite{Sutula:2001}.  Many theoretical and empirical correlations and computational fire models have been developed for use in fire protection design, fire hazard analysis, and fire investigation.  A collection of the widely used correlations can be found in the U.S. Nuclear Regulatory Commission’s (NRC) Fire Dynamics Tools (FDTs)~\cite{NUREG_1805}. These spreadsheets incorporate references to the source documents and experimental work upon which the correlations are based. NRC tasked NIST with conducting a verification and validation study of these correlations. Overholt developed the NIST verification and validation study~\cite{Overholt:2014}. Not all of the correlations in the FDTs were included in the study, because a sufficient amount of reliable data was not available to conduct a proper validation.  The correlations for flame height and time to ignition were not included in the validation study.

@@

@@

The majority of intentional fires in structures in the United States have occurred in residential occupancies, so this study focused on fire patterns on painted gypsum wallboard, which is a common interior finish for residential occupancies.  A common height for a residential occupancies is 2.4~m.  With a wall height of 2.4~m, these experiments used small fires with flame heights of approximately 1.2~m.  The resulting patterns on the gypsum board wall were be representative of a pre-flashover fire pattern.

In order to understand the variability or repeatability of fire patterns, the repeatability of the fire generating the fire pattern must be quantified.  The fires were characterized in terms of heat release rate, mass loss rate (when appropriate), plume temperature, heat flux, and visible flame height.

@@

@@

Cone calorimeter experiments were conducted, based on ASTM~E 1354, to estimate the ignition temperature, the time to ignition, and the heat release rate from samples of painted gypsum wallboard under thermal flux conditions ranging from 25~kW/m$^2$ to 75~kW/m$^2$ ~\cite{ASTM_E1354}.

\subsection{Numerical Fire Pattern Analysis}

The ability to recreate or simulate a fire pattern or certain characteristics of a fire pattern was examined with a variety of tools.  The first characteristic examined was the ability to determine, based on the calculated heat flux to the wall, if a visible or measurable physical change would occur to the painted paper surface of the gypsum wallboard.  The next characteristic of the fire pattern examined was the height.  The height of the fire pattern was then compared to the predictive results from flame height correlations. Heskestad developed an empirical method for calculating the mean visible flame height for fuels that do not have substantial in-depth combustion~\cite{Beyler:1986,Heskestad:SFPE}. This correlation was compared with the flame height results of the source fires and the instrumented wall experiments.

@@

The chair has more unburned foam cushioning then the sofa does, and the seat cushion is still in place with it's distinctive shape, so you know that cushion has not been moved.  The cushion was heated from the top down.  Perhaps the construction of the chair slowed the flame spread through the chair from the bottom up, if the heater was the ignition source.  Or the fire started in the sofa and the radiation from the flames and hot gas layer spread the fire to the chair.

@@

Flexible polyurethane foam was chosen as the solid fuel. According to the Polyurethane Foam Association, flexible polyurethane foam is the most widely used cushioning material in upholstered furnishings and mattresses. More than 544 million kg are produced in the U.S. every year~\cite{Foam,Polyurethane_Foam}. In studies by Ahrens and Evarts of the NFPA, upholstered furniture, and mattresses and bedding were the first items ignited in 33\% of the fatal fires in the U.S. based on averaged data from 2005 through 2009~\cite{Ahrens:2011,Evarts:2011}.

@@

@@

The calculations for the intensity field are based on a user-defined computation grid.  A grid sensitivity analysis was conducted with 5~mm, 10~mm, 20~mm, 40~mm, and 80~mm square grids.  Examples of grid cell resolution outputs from Streams for the natural gas fueled burner are shown in Fig.~\ref{Intensity}.  The outputs shown are time-averaged intensity fields generated with the resolutions listed above. In each output the x and y axis dimensions are in millimeters.

\begin{figure}[p]
	\centering
	\includegraphics[trim=0in 0in 3.5in 0in,clip=true,width=1.7in]{../Figures/FHNG80_1_GS_5mm_10s_color}
	\includegraphics[trim=0in 0in 3.5in 0in,clip=true,width=1.7in]{../Figures/FHNG80_1_GS_10mm_10s_color}\\
	\includegraphics[trim=0in 0in 3.5in 0in,clip=true,width=1.7in]{../Figures/FHNG80_1_GS_20mm_10s_color}
	\includegraphics[trim=0in 0in 3.5in 0in,clip=true,width=1.7in]{../Figures/FHNG80_1_GS_40mm_10s_color}
	\includegraphics[trim=0in 0in 3.5in 0in,clip=true,width=1.7in]{../Figures/FHNG80_1_GS_80mm_10s_color}\\
	 \caption[Comparison images of intensity fields derived from the video images of the natural gas fueled fire]{Comparison images of intensity fields derived from the video images of the natural gas fueled fire using Streams. The images, starting with the upper left, were generated with 5~mm, 10~mm, 20~mm, 40~mm, and 80~mm resolutions.}
	 \label{Intensity}
\end{figure}

@@

\begin{figure}
  \includegraphics[width=2.7in]{../Figures/Fig20}
  \includegraphics[width=2.7in]{../Figures/Fig21} \\
  \caption[Visibility fractions for the natural gas and gasoline fires]{Plots showing the fraction of time the flame was visible for the natural gas (left) and gasoline (right) fires.}
  \label{Gas_and_Gasoline_Contours}
\end{figure}

@@

\subsection{Natural Gas Burner Plume and Wall Exposures}
Figures~\ref{TC_Plume_TWNG_comp} through \ref{TC_Back_Center_TWNG_comp} show a complete set of data, averaged over two replicates, for the instrumented, non-combustible wall, natural gas burner experiments.  Comparing the temperatures of the plume at the most distant point from the wall (0.61~m) and with the plume temperatures with the burner contacting the wall in Figure~\ref{TC_Plume_TWNG_comp} shows an increase.  However, for the burner position 0.15~m from the wall, a decrease in plume temperatures occurs approximately 3~min after ignition.  It is worth noting that as the plume temperatures decrease, the temperatures of the thermocouples installed in the wall face increased, indicating that after the wall heats up the flame pulls toward the wall.  Perhaps as a result of increased convection on the face of the wall.

@@

The physics of the algorithm are sound.  The challenge for the end user is determining the appropriate input values and making sure to use the algorithm within the appropriate distance range.  More information or research is needed on the radiative fractions of various fuels.  For example, the majority of natural gas is composed of methane.  However, natural gas may also contain small amounts of ethane, propane, hexane, butane, and pentane. Some of these gases have higher radiative fractions than methane. In one widely used text, a table listing radiative energy fraction values for burner diameters larger than 0.5~m lists butane in the 20 to 40 $\%$ range and hexane in the 40 to 60 $\%$ range ~\cite{Quintiere:1997}.  So, it is possible that the radiative fraction for natural gas could be higher that the radiative fraction associated with methane.  The radiative fraction value in the literature for gasoline is 0.45~\cite{Quintiere:1997, IAAI:2017}.  This is in line with the determined value.

@@

The FDS images are compared with three photographs of fire patterns from the compartment fire pattern experiments documented in Chapter~6.  The three images from left to right exhibit a range of fire pattern heights from approximately 0.7 to 0.8~m.  The average height of the natural gas fire patterns from this open wall construction series was 0.74 with a 95\% certainty range of $\pm$0.12~m. The simulated fire pattern was taller than the actual patterns.  However if the upper ignition temperature was used, the height values between the actual and the simulated would have been similar.  The RI=10 simulation was used as a means to optimize the graphical resolution, even though the temperature range of the flames were shifted up relative to the RI=5 case.

@@

A few general thoughts to place the numerical methods in perspective.  In the experimental portion of this study, we established with the gas burner experiments that the steady state heat release rate values had a 95\% certainty range of $\pm~11\%$.  This range was the same as that associated with the heat release rate measurement system, so this would be the best agreement we should expect.  The gasoline and the polyurethane introduced additional variability, and the 95\% certainty ranges on the peak heat release rate values increased to 30\% and 47\%, respectively.  As the physical variability of the experiments increases, what is the expectation of numerical methods that simulate the physical experiments?

@@

In Chapter 1, as the investigator you had completed your review of the scene and had collected evidence, which included an electric heater.  You are not sure about the fire's progression or the area of origin.  Before leaving the scene, you meet up with the fire captain and begin asking him about the fire.  The captain tells you the doors and windows of the townhouse were closed upon arrival.  When the front door was opened, heavy smoke flowed out, followed by flames.  Other fire fighters were venting the upstairs windows.  The fire grew and burned for a short time before water was applied through the front door and the fire was knocked down.

Comparing the captain's description with your scene, it would seem that the fire started and burned as a fuel-limited fire, before transitioning to a ventilation-limited 
fire in which the heat and fuel where adequate to sustain combustion, but perhaps oxygen was in short supply.  This fits the description of the fire flaring up and exiting the front door shortly after it was opened.  With the information that other firefighters were opening the upstairs windows at this time, you starting thinking about the flow path in the town home.  The front door served as a bi-directional vent with the lower portion serving as an air inlet and the upper portion serving as an exhaust for flames.  In the upstairs bedroom there was mostly smoke damage and no significant heat damage. It seems the open bedroom window served as a smoke exhaust vent.

The large V pattern on the wall behind the sofa and the stair wall may have been generated after the door and upstairs window was opened. This is an area where the fuel-rich hot fire gases would mix with fresh air entering the front door and burn.  The area of the living room more remote from the front door, (the source of oxygen), did not have as much fire damage.  You remember there was fire damage to the wall behind the right side of the sofa.  This burned area on the wall seems distinct and disconnected from the area with the most fire damage, which is to the left of the sofa.

You decide to go back in the town house for another look Figure~\ref{Pre_Flashover_Pattern}.  The pattern behind the sofa had to be generated early in the fire event because it is away from the front door. Further, the wall behind the sofa between this pattern and the larger area of fire damage closer to the front door was protected and not exposed to flaming combustion.  The inside of the sofa's seat frame was charred on the inside but not on the outside in the area in front of this pattern, Figure~\ref{Table}.  The bottom of the pattern is located near the base of the sofa seat cushion.  The pattern height is approximately 0.7~m high extending just above the top of the wood sofa frame.  The base of the pattern is approximately 0.3~m.

Using Heskestad's flame height equation and solving for the heat release rate based on the pattern height yields a heat release rate on the order of 55~kW. Unknowns include the impact of the back portion of the sofa, but this fire size seems reasonable to you with regard to the vision of a sofa cushion burning. This information, coupled with the sequence of events provided by the fire captain and other fire damage, indicate the right side of the sofa is likely the area of origin even though it does not have the most fire damage.  You still need to interview the occupants.  Was an accidental ignition due to careless smoking possible? Or a lithium ion battery powered device?  Also, you will still check if the heater switch was in the on position.  If not, that further supports your current hypothesis about the right side of the sofa.  You will also take samples of the debris in the area of the right seat cushion to send to the lab.  You have done all that you can at the scene.

If needed, you have the dimensions of the house and the opening so you could have a computer model of the house made to examine how long it would take a small fire to decrease the level of the oxygen within the house to 15\%.  This might prove necessary to get insight on a time line you might get from the interviews of the occupants.

\begin{figure}[h]
	\includegraphics[width=5in]{../Figures/Pre_Flashover_Pattern} \\
	\caption[Photograph showing the wall with the stand alone fire pattern toward the right side of the sofa.]{Photograph showing the wall with the standalone fire pattern toward the right side of the sofa.}
	\label{Pre_Flashover_Pattern}
\end{figure}

@@

N.B.: This section was included to provide a sense of how a fire investigator might use their knowledge of fire behavior, fire patterns, and numerical methods to assist them in following the scientific method.  Fire investigation requires significantly more than pattern interpretation.  The totality of facts from the whole situation must gathered to the extent possible, and then after analysis of the evidence, a fire cause may be determined.

@@

@@

There is a great need to understand the impact of ventilation on fire patterns within a structure.  Fires that stop pre-flashover and fires that burn post-flashover need study.  The structure itself will have an impact on the movement and flow of both fresh air to the fire and containing and channeling the movement of combustion products away from the fire, both of which can affect the development of the fire pattern.  Post-flashover fires are particularly challenging for fire investigators because the potential is high significant fire damage may be generated remote from the area of origin.

@@

The main objective of this research study was to examine the repeatability of pre-flashover fire patterns on gypsum wallboard.  The patterns were generated from exposure to short duration (300 seconds maximum), well characterized fires.

The first step was to examine the repeatability of different fire sources.  Three different fuels were used: natural gas, gasoline, and polyurethane foam.  Each fuel source had a similar top surface area.  The heat release rate data showed that the variability was greater for more complex fuels.  The variation in the average peak heat release rate with the natural gas was similar to the expanded uncertainty of the measurement system, $\pm$ 11\%.  However, the variation in the average peak heat release rates of the gasoline and the polyurethane foam increased due to uncertainties in the burning behavior of the fuels.  The average peak heat release rate of gasoline had a 95\% confidence limit of $\pm$~30\% and the same value for the polyurethane foam was $\pm$~47\%.

The fire pattern heights generated on painted gypsum wallboard by the natural gas burner and gasoline fueled fires were shown to have uncertainties of $\pm$ 18\% or less based on a Type A statistical analysis with 95\% confidence limits. The patterns generated by the polyurethane foam fueled fire had more than twice the variability of the natural gas and the gasoline fire patterns.  This was likely due to the greater variability of the burning of the solid fuel and the lower heat release rate.

>>>>

In the United States, the loss of life due to unwanted fire has been decreasing steadily since the 1970’s.  (This trend does not include loss of life as a result of the terrorist attack on September 11, 2001.)  The estimated number of fire deaths in the U.S. in 1975 was 8,100, or 3.75 deaths per 100,000 people~\cite{America_Burning_Revisited}.  In 2013, there were 3,240 civilian fire fatalities, or 1.02 deaths per 100,000 people~\cite{Karter:2014}.  During this same period, the number of fires in the United States has decreased from 2.5 million fires to 1.2 million fires.

There are many reasons that led to this decrease in loss of life due to fire in the United States.  One example is research and data that enabled the development of product safety standards, improved building codes, improved fire safety systems, and fire prevention programs.  None of this happened without the constant effort of agencies at the federal, state, and local levels working with professional fire organizations in order to deliver the five E’s of fire prevention: engineering, education (of the public), enforcement, economic incentive, and emergency response~\cite{FEMA:2013}.

Fire investigation is one element in the fire prevention system that generates data to enable the reduction of fire losses.  The investigation of fires provides a means to identify the cause of the fire in order to develop a knowledge base that could enable the elimination of that cause, and thus further reduce the losses from fires.  Data such as the room of fire origin, the first item ignited, and ultimately the cause of the fire is information that is critical to understanding and reducing the number of fires.  For example, the identification of products that are involved in ignitions due to a design flaw in many cases result in the U.S. Consumer Product Safety Commission or a manufacturer issuing a product recall. There are two notable and recent examples of this process. As result of reports of fire and overheating incidents, 1.9 million Samsung Galaxy Note7 Smartphones and 0.5 million hoverboards were recalled in 2016~\cite{CPSC:Samsung,CPSC:Hoverboards}.

@@

\subsection{The Use of Fire Models in Fire Investigation}
Fire models are another tool that may be used by an engineer on a fire investigation team to gain insight into the growth and development of a fire and to test the investigators' hypothesis on origin and cause~\cite{Sutula:2001}.  Many theoretical and empirical correlations and computational fire models have been developed for use in fire protection design, fire hazard analysis, and fire investigation.  A collection of the widely used correlations can be found in the U.S. Nuclear Regulatory Commission’s (NRC) Fire Dynamics Tools (FDTs)~\cite{NUREG_1805}. These spreadsheets incorporate references to the source documents and experimental work upon which the correlations are based. NRC tasked NIST with conducting a verification and validation study of these correlations. Overholt developed the NIST verification and validation study~\cite{Overholt:2014}. Not all of the correlations in the FDTs were included in the study, because a sufficient amount of reliable data was not available to conduct a proper validation.  The correlations for flame height and time to ignition were not included in the validation study.

@@

Despite the current validation studies, the implementation of fire models for a room and contents fire in a home is challenging for a number of reasons which include; the lack of appropriate input data, the differences between well characterized fuels used for validation studies and the fuels found in homes, ventilation limited fire behavior, and the inability to reliably predict failure of windows or wall systems (i.e. changes in ventilation).

@@

The topic of fire pattern repeatability covers scenarios that could range from a small kitchen fire to a multi-compartment, multi-story fire in a commercial building.  Therefore the scope of study had to be narrowed considerably.

The majority of intentional fires in structures in the United States have occurred in residential occupancies, so this study focused on fire patterns on painted gypsum wallboard, which is a common interior finish for residential occupancies.  A common height for a residential occupancies is 2.4~m.  With a wall height of 2.4~m, these experiments designed for small fires with flame heights of approximately 1.2~m in order to allow for additional vertical flame spread on the wall, without exceeding the height of the wall. The resulting patterns on the gypsum board wall were representative of a pre-flashover fire pattern.

In order to understand the variability or repeatability of fire patterns, the repeatability of the fire generating the fire pattern must be quantified.  The fires were characterized in terms of heat release rate, mass loss rate (when appropriate), plume temperature, heat flux, and visible flame height. While conducting scoping studies the size of the burner or fuel surface area was determined.

One parameter that needed to be considered was time.  These fires were intended to represent the generation of a pre-flashover burn pattern in a residential scale room. Studies conducted in residential scale structures with furnished rooms showed that the fire development in the room of origin was reduced due to ventilation limits within 5 minutes or less after ignition of a bed or upholstered furniture item with a flaming ignition source~\cite{Kerber:2010,Madrzykowski:2009}. Even in cases where the compartment transitioned through flashover, the fire in the area of origin may have stopped burning due to a localized lack of oxygen. A burn time of 300 seconds was chosen.

Time is not a critical factor when using a natural gas fueled burner.  However, the 300 second limit was also a practical consideration to ensure that enough gasoline or polyurethane foam could be repeatably and reliably supplied to burn for the duration of a given experiment.

@@

The critical heat flux boundary has been defined in full scale experiments as the limit of fire propagation by Newman and Tewarson~\cite{Newman}. This concept has also been identifed in bench scale radiant panel tests~\cite{Adams:1977}.

@@

Cone calorimeter experiments were conducted, based on ASTM~E 1354, to estimate the ignition temperature, the time to ignition, and the heat release rate from samples of painted gypsum wallboard under thermal flux conditions ranging from 25~kW/m$^2$ to 75~kW/m$^2$ ~\cite{ASTM_E1354}.
{}
\subsection{Numerical Fire Pattern Analysis}
{}
The ability to recreate or simulate a fire pattern or certain characteristics of a fire pattern was examined with a variety of tools.  The first characteristic examined was the ability to determine, based on the calculated heat flux to the wall, if a visible or measurable physical change would occur to the painted paper surface of the gypsum wallboard.  The next characteristic of the fire pattern examined was the height.  The height of the fire pattern was then compared to the predictive results from flame height correlations. Heskestad developed an empirical method for calculating the mean visible flame height for fuels that do not have substantial in-depth combustion~\cite{Beyler:1986,Heskestad:SFPE}. This correlation was compared with the flame height results of the source fires and the instrumented wall experiments.

@@

The chair has more unburned foam cushioning then the sofa does, and the seat cushion is still in place with its distinctive shape, so you know that cushion has not been moved.  The cushion was heated from the top down.  Perhaps the construction of the chair slowed the flame spread through the chair from the bottom up, if the heater was the ignition source.  Or the fire started in the sofa and the radiation from the flames and hot gas layer spread the fire to the chair.

@@

Flexible polyurethane foam was chosen as the solid fuel. According to the Polyurethane Foam Association, flexible polyurethane foam is the most widely used cushioning material in upholstered furnishings and mattresses. More than 544 million kg are produced in the U.S. every year~\cite{Foam,Polyurethane_Foam}. In studies by Ahrens and Evarts of the NFPA, upholstered furniture, and mattresses and bedding were the first items ignited in 33\% of the fatal fires in the U.S. based on averaged data from 2005 through 2009~\cite{Ahrens:2011,Evarts:2011}.  Between 2007 and 2011, 11\% of all of the mattress and bedding fires and 9\% of all of the upholstered furniture fires in the United states were ignited by smoking materials.  Fires ignited with smoking material are a leading cause of fatal fires, with 29\% and 31\% of all fatal fires involving mattresses and bedding, and upholstered furniture, respectively~\cite{Hall:2013}.  For this study, the polyurethane foam would be ignited with a small flame to reduce some of the uncertainties associated with a smoldering cigarette ignition.

@@

For several of the source fire characterization fires, an additional Schmidt-Boelter total heat flux gauges was positioned 2~m located 0.95~m horizontally from the center of the burner/fuel and 0.50~m above the top of the burner/pan lip/fuel surface.  These 3 gauges were used to measure the heat flux to determine the radiative fraction for each fuel.

@@

The calculations for the time-averaged cumulative light intensity field are based on a user-defined computation grid.  A grid sensitivity analysis was conducted with 5~mm, 10~mm, 20~mm, 40~mm, and 80~mm square grids.  Examples of grid cell resolution outputs from Streams for the natural gas fueled burner are shown in Fig.~\ref{Intensity}.  The outputs shown are time-averaged light intensity fields generated with the resolutions listed above. In each output the x and y axis dimensions are in millimeters.

\begin{figure}[p]
	\centering
	\includegraphics[trim=0in 0in 3.5in 0in,clip=true,width=1.7in]{../Figures/FHNG80_1_GS_5mm_10s_color}
	\includegraphics[trim=0in 0in 3.5in 0in,clip=true,width=1.7in]{../Figures/FHNG80_1_GS_10mm_10s_color}\\
	\includegraphics[trim=0in 0in 3.5in 0in,clip=true,width=1.7in]{../Figures/FHNG80_1_GS_20mm_10s_color}
	\includegraphics[trim=0in 0in 3.5in 0in,clip=true,width=1.7in]{../Figures/FHNG80_1_GS_40mm_10s_color}
	\includegraphics[trim=0in 0in 3.5in 0in,clip=true,width=1.7in]{../Figures/FHNG80_1_GS_80mm_10s_color}\\
	 \caption[Comparison images of time-averaged light intensity fields derived from the video images of the natural gas fueled fire]{Comparison images of time-averaged light intensity fields derived from the video images of the natural gas fueled fire using Streams. The images, starting with the upper left, were generated with 5~mm, 10~mm, 20~mm, 40~mm, and 80~mm resolutions.}
	 \label{Intensity}
\end{figure}

@@

\begin{figure}
  \includegraphics[width=2.42in]{../Figures/Fig20}
  \includegraphics[width=2.7in]{../Figures/Fig21} \\
  \caption[Visibility fractions for the natural gas and gasoline fires]{Plots showing the fraction of time the flame was visible for the natural gas (left) and gasoline (right) fires.}
  \label{Gas_and_Gasoline_Contours}
\end{figure}

@@

\subsection{Natural Gas Burner Plume and Wall Exposures}
Figures~\ref{TC_Plume_TWNG_comp} through \ref{TC_Back_Center_TWNG_comp} show a complete set of data, averaged over two replicates, for the instrumented, non-combustible wall, natural gas burner experiments.  Comparing the temperatures of the plume at the most distant point from the wall (0.61~m) and with the plume temperatures with the burner contacting the wall in Figure~\ref{TC_Plume_TWNG_comp} shows an increase.  However, for the burner position 0.15~m from the wall, a decrease in plume temperatures occurs approximately 3~min after ignition.  It is worth noting that as the plume temperatures decrease, the temperatures of the thermocouples installed in the wall face increased, indicating that after the wall heats up the flame pulls toward the wall.  Perhaps as a result of increased convection on the face of the wall.  the distance that the flame moved toward the wall varied.  At 0.15~m distance from the wall, after 150 to 200~ seconds, the flame would intermittently contact the wall.

@@

The physics of the algorithm are sound.  The challenge for the end user is determining the appropriate input values and making sure to use the algorithm within the appropriate distance range.  More information or research is needed on the radiative fractions of various fuels.  For example, the majority of natural gas is composed of methane.  However, natural gas may also contain small amounts of ethane, propane, hexane, butane, and pentane. Some of these gases have higher radiative fractions than methane. In one widely used text, a table listing radiative energy fraction values for burner diameters larger than 0.5~m lists butane in the 20 to 40 $\%$ range and hexane in the 40 to 60 $\%$ range ~\cite{Quintiere:1997}.  So, it is possible that the radiative fraction for natural gas could be higher than the radiative fraction associated with methane.  The radiative fraction value in the literature for gasoline is 0.45~\cite{Quintiere:1997, IAAI:2017}.  This is in line with the determined value.

@@

The FDS images are compared with three photographs of fire patterns from the compartment fire pattern experiments documented in Chapter~6.  The three images from left to right exhibit a range of fire pattern heights from approximately 0.7 to 0.8~m.  The average height of the natural gas fire patterns from this open wall construction series was 0.74 with a 95\% uncertainty range of $\pm$0.12~m. The simulated fire pattern was taller than the actual patterns.  However if the upper ignition temperature was used, the height values between the actual and the simulated would have been similar.  The RI=10 simulation was used as a means to optimize the graphical resolution, even though the temperature range of the flames were shifted up relative to the RI=5 case.

@@

A few general thoughts to place the numerical methods in perspective.  In the experimental portion of this study, we established with the gas burner experiments that the steady state heat release rate values had a 95\% uncertainty range of $\pm~11\%$.  This range was the same as that associated with the heat release rate measurement system, so this would be the best agreement we should expect.  The gasoline and the polyurethane introduced additional variability, and the 95\% uncertainty ranges on the peak heat release rate values increased to 30\% and 47\%, respectively.  As the physical variability of the experiments increases, what is the expectation of numerical methods that simulate the physical experiments?

In this study, a single oxygen consumption calorimeter was used to measure the HRRs of the source fuels used in this study.  This calorimeter has been associated with a validation study and was calibrated throughout the study~\cite{Bryant:2003b}.  In 2000, Janssens conducted a round robin study on the repeatability of heat release rates from full scale oxygen consumption furniture calorimeters.  The study involved 4 laboratories, each burning 3 replicates of 4 different fuels. The range of relative repeatability standard deviations were 16 to 58\% and 34 to 69\% for peak HRR and total heat released, respectively~\cite{Janssens:2002}.  The lack of uniformity of the results from these measurement devices is of concern and is clearly another area that needs to be researched and addressed, especially if a single organization is going to be generating HRR data for database or modeling input purposes.

@@

In Chapter 1, as the investigator you had completed your review of the scene and had collected evidence, which included an electric heater.  At this point, you are not sure about the fire's progression or the area of origin.  Before leaving the scene, you meet up with the fire captain and begin asking him about the fire.  The captain tells you the doors and windows of the townhouse were closed upon arrival.  When the front door was opened, heavy smoke flowed out, followed by flames.  Other fire fighters were venting the upstairs windows on the front of the townhouse.  The fire grew and burned for a short time before water was applied through the front door and the fire was knocked down.

\subsection{Hypothesis Testing, Applying Fire Dynamics}

Comparing the captain's description with your scene and applying your knowledge of fire dynamics, it would seem that the fire started and burned as a fuel-limited fire, before transitioning to a ventilation-limited fire in which the heat and fuel where adequate to sustain combustion, but perhaps oxygen was in short supply.  This fits the description of the fire flaring up and exiting the front door shortly after it was opened.  With the information that other firefighters were opening the upstairs windows at this time, you begin thinking about the flow path in the town home.  The front door served as a bi-directional vent with the lower portion serving as the air (oxygen) inlet and the upper portion serving as an exhaust for flames and combustion products.  In the upstairs bedroom there was mostly smoke damage and no significant heat damage. It seems the open bedroom window served as a smoke exhaust vent.

The large V pattern on the corner behind the chair next to the stair wall may have been generated post-flashover, after the flow paths between the front door and upstairs window were opened. This is an area where the fuel-rich hot fire gases would mix with fresh air entering the front door and burn. In other words, the basic qualitative requirements of the fire triangle could be met.  The area of the living room more remote from the front door (the source of oxygen), did not have as much fire damage.  You remember there was limited fire damage to the wall behind the right side of the sofa.  This burned area on the wall seemed distinct and disconnected from the area with the most fire damage, which is to the left of the sofa.

You decide to go back into the town house for another look Figure~\ref{Pre_Flashover_Markup}.  The dark colored fire pattern (inside yellow oval), behind the sofa and above the location of the right seat cushion, had to be generated early in the fire event because it is away from the front door. Further, the wall behind the sofa between this pattern and the larger area of fire damage (V pattern between red lines in the left corner) closer to the front door was protected and not exposed to flaming combustion.  Comparing the views of the sofa frame from Figure~\ref{Pre_Flashover_Markup} and Figure~\ref{Table}, it can be seen that the inside of the sofa's seat frame was charred on the inside but not on the outside in the area in front of this pattern.  The burning inside the sofa frame may have been the result of the sofa cushion melting, dripping, burning.  The bottom of the pattern is located near the base of the sofa seat cushion.  The pattern height is approximately 0.7~m high extending just above the top of the wood sofa frame.  The base of the pattern is approximately 0.3~m.

Using Heskestad's flame height equation and solving for the heat release rate based on the pattern height yields a heat release rate on the order of 55~kW. There are other unknowns (for example the fire contribution of the back portion of the sofa, the impact of the melting, dripping, and burning seat cushion) but this fire size seems reasonable with regard to the vision of a sofa cushion burning early in the fire incident. This information, coupled with the sequence of events provided by the fire captain and other fire damage, indicate the right side of the sofa is likely the area of origin even though it does not have the most fire damage.  You still need to interview the occupants, to get additional information to examine your hypothesis.  Was an accidental ignition due to careless smoking possible? Or a lithium ion battery powered device?  Also, you will still want a forensic analysis of the heater switch, was in the on position?  If not, that further supports your current hypothesis about the right side of the sofa.  You will also take samples of the debris in the area of the right seat cushion to send to the lab.  You have done all that you can at the scene.

If needed, you have taken the dimensions of the house and the openings so you could have a computer model developed to examine how long it would take a small fire to decrease the level of the oxygen within the house to below 15\%.  This might prove necessary to get insight on the time line you might get from the interviews of the occupants.

\begin{figure}[h]
	\includegraphics[width=5in]{../Figures/Pre_Flashover_Markup} \\
	\caption[Photograph showing the wall with the stand alone fire pattern toward the right side of the sofa.]{Photograph showing the wall with the stand alone fire pattern toward the right side of the sofa (inside yellow oval). Note that the wall behind the sofa and to the left of the stand alone fire pattern does not have significant fire damage. While, the corner area to the left of the sofa and behind the chair has a V pattern shaped fire pattern}
	\label{Pre_Flashover_Markup}
\end{figure}

@@

N.B.: This section was included to provide a sense of how a fire investigator might use their knowledge of fire behavior, fire patterns, and numerical methods to assist them in following the scientific method.  Fire investigation requires significantly more analysis than pattern interpretation.  The totality of facts from the whole situation must gathered to the extent possible, and then after analysis of the evidence, an area of origin can be located and the fire cause may be determined.  In some cases, despite a trained investigators best efforts, the cause of the fire may have to be documented as "undetermined".  In Chapter 1, information was provided about the occupants of the house.  The investigator must be careful that information about the occupants, or other information not related to evidence at the scene, does not create a cognitive bias.

The photographs used to support this scenario were taken from a fire experiment in an acquired structure.  The fire was ignited on the right side of the sofa with a small flaming source.  The ventilation of the structure occurred as described and resulting in the fire patterns as shown.

An improved understanding of the cause and effect relationship between the heat release rate, heat transfer, and the resulting fire damage on simple targets, such as walls, can be used to assess fire damage such as the stand alone, preflashover pattern in this scenario.  Further validation and an improved understanding of both the capabilities and limitations of numerical methods to recreate fire patterns is needed to assist in hypothesis testing.

This study provides a foundational understanding of the repeatability of pre-flashover patterns, such as the one shown in this fire Investigator's Challenge.  While the existing numerical methods have been examined with simple source fires in this study and others, further examination is needed with typical first items ignitied, such as a upholstered chair or sofa to understand the applicability and the impact of the compartment.

@@

A oxygen consumption calorimeter round robin study should be conducted among the major federal and private laboratories to determine if the conditions that Janssens reported in 2000 still exist, and if so, what is the underlying cause for the measurement discrepancies~\cite{Janssens:2002}. This type of study is critical for both the fundamental understanding of the fire behavior of materials as well as the need for quality input data for numerical models.

@@

There is a great need to understand the impact of ventilation on fire patterns within a structure.  Fires that stop pre-flashover and fires that burn post-flashover need study.  The structure itself will have an impact on the movement and flow of both fresh air to the fire and containing and channeling the movement of combustion products away from the fire, both of which can affect the development of the fire pattern.  Post-flashover fires are particularly challenging for fire investigators because the potential is high that significant fire damage may be generated remote from the area of origin.

@@

The main objective of this research study was to examine the repeatability of pre-flashover fire patterns on gypsum wallboard.  The patterns were generated from exposure to short duration (300 seconds maximum), well characterized fires.  The results provide investigators with baseline data on the repeatability of fire patterns under near ideal conditions.  Applications of analyzing the patterns with existing data from the literature, as well as with a range of numerical methods, demonstrated a number of current limitations and the need for additional research and data to support the application of engineering analysis of fire patterns.

The first step was to examine the repeatability of different fire sources.  Three different fuels were used: natural gas, gasoline, and polyurethane foam.  Each fuel source had a similar top surface area.  The heat release rate data showed that the variability was greater for more complex fuels.  The variation in the average peak heat release rate with the natural gas was similar to the expanded uncertainty of the measurement system, $\pm$ 11\%.  However, the variation in the average peak heat release rates of the gasoline and the polyurethane foam increased due to uncertainties in the burning behavior of the fuels.  The average peak heat release rate of gasoline had a 95\% confidence limit of $\pm$~30\% and the same value for the polyurethane foam was $\pm$~47\%.  Therefore applying the HRR uncertainty of the measurement device to numerical method input values is not sufficient, the uncertainty related to the behavior or composition of the fuel must also be taken into consideration.
